---
title: "Introduction to metavizr: interactive visualization for metagenomics"
author: "Héctor Corrada Bravo, Florin Chelaru, Justin Wagner, Jayaram Kancherla, Joseph Paulson"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to metavizr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[`Metaviz`](http://metaviz.cbcb.umd.edu) is an interactive visualization tool for metagenomic sequencing data. It supports taxonomy navigation of annotated metagenomic features and allows multiple visualizations of data scatterplots, heatmaps and PCA plots. It also includes data from the [Mild to Severe Diahereal Study](http://msd16s/) for visualization. As with the Epiviz suite, it has a flexible plugin framework so users can add [d3](http://d3js.org/) visualizations. You can find more information about Metaviz at [http://epivizwww01.umiacs.umd.edu/help](http://epivizwww01.umiacs.umd.edu/help) and see a video tour.

<iframe width="480" height="360" src="http://www.youtube.com/embed/099c4wUxozA" frameborder="1" allowfullscreen></iframe>

The `metavizr` package implements two-way communication between the `R/Bioconductor` computational genomics environment and `Metaviz`. Objects in an `R` session can be displayed as tracks or plots on Metaviz Metavizr uses Websockets for communication between the browser Javascript client and the R environmen, the same technology underlying the popular [Shiny](http://www.rstudio.com/shiny/) system for authoring interactive web-based reports in R.

### Preliminaries: the data

In this vignette we will look at the mouse data from the `metagenomeSeq` package.

```{r, eval=TRUE, echo=TRUE, results='hide', warning=FALSE, error=FALSE}
require(metavizr)
require(metagenomeSeq)

data("mouseData")
```

### The metavizr session manager

The connection to `Metaviz` is managed through a session manager object of class `EpivizApp`. We can create this object and open `Metaviz` using the `startMetaviz` function.

```{r, eval=FALSE, echo=TRUE}
app <- startMetaviz()
```

```{r, eval=TRUE, echo=FALSE}

app <- startMetaviz(host="http://localhost", http_port=8989, debug=TRUE, open_browser=FALSE, non_interactive=TRUE, try_ports=TRUE)

# register IciclePlot

js_chart_settings <- list(list(id = "title", type = "string", defaultValue = "", label = "Title", possibleValues = NULL), list(id = "marginTop", type = "number", defaultValue = 50, label = "Top margin", possibleValues = NULL), list(id = "marginBottom", type = "number", defaultValue = 23, label = "Bottom margin", possibleValues = NULL), list(id = "marginLeft", type = "number", defaultValue = 20, label = "Left margin", possibleValues = NULL), list(id = "marginRight", type = "number", defaultValue = 10, label = "Right margin", possibleValues = NULL))

js_chart_colors = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf")

app$chart_mgr$register_chart_type("epiviz.ui.charts.tree.Icicle", "epiviz.ui.charts.tree.Icicle", js_chart_settings=js_chart_settings, js_chart_colors=js_chart_colors)


# register HeatpmapPlot

js_chart_settings <- list(list("id"="title","type"="string","defaultValue"="","label"="Title","possibleValues"=NULL),list("id"="marginTop","type"="number","defaultValue"=80,"label"="Top margin","possibleValues"=NULL),list("id"="marginBottom","type"="number","defaultValue"=40,"label"="Bottom margin","possibleValues"=NULL),list("id"="marginLeft","type"="number","defaultValue"=120,"label"="Left margin","possibleValues"=NULL),list("id"="marginRight","type"="number","defaultValue"=40,"label"="Right margin","possibleValues"=NULL),list("id"="measurementGroupsAggregator","type"="categorical","defaultValue"="mean-stdev","label"="Aggregator for measurement groups","possibleValues"=list("mean-stdev","quartiles","count","min","max","sum")),list("id"="colLabel","type"="measurementsMetadata","defaultValue"="colLabel","label"="Columns labels","possibleValues"=NULL),list("id"="rowLabel","type"="measurementsAnnotation","defaultValue"="name","label"="Row labels","possibleValues"=NULL),list("id"="showColorsForRowLabels","type"="boolean","defaultValue"=FALSE,"label"="Row labels as colors","possibleValues"=NULL),list("id"="maxColumns","type"="number","defaultValue"=40,"label"="Max columns","possibleValues"=NULL),list("id"="yMin","type"="number","defaultValue"="default","label"="Min Value","possibleValues"=NULL),list("id"="yMax","type"="number","defaultValue"="default","label"="Max Value","possibleValues"=NULL),list("id"="cluster","type"="categorical","defaultValue"="rows","label"="Cluster","possibleValues"=list("none","rows","columns","both")),list("id"="clusteringAlg","type"="categorical","defaultValue"="none","label"="Clustering Algorithm","possibleValues"=list("none","agglomerative")),list("id"="clusteringMetric","type"="categorical","defaultValue"="euclidean","label"="Clustering Metric","possibleValues"=list("euclidean")),list("id"="clusteringLinkage","type"="categorical","defaultValue"="complete","label"="Clustering Linkage","possibleValues"=list("complete")),list("id"="showDendrogram","type"="boolean","defaultValue"=TRUE,"label"="Show Dendrogram","possibleValues"=NULL))

js_chart_colors = c("#1859a9","#ed2d2e","#008c47","#010101","#f37d22","#662c91","#a11d20","#b33893")

app$chart_mgr$register_chart_type("HeatmapPlot", "epiviz.plugins.charts.HeatmapPlot", js_chart_settings=js_chart_settings, js_chart_colors=js_chart_colors)


# register StackedLinePlot

js_chart_settings <- list(list("id"="title","type"="string","defaultValue"="","label"="Title","possibleValues"=NULL),list("id"="marginTop","type"="number","defaultValue"=30,"label"="Top margin","possibleValues"=NULL),list("id"="marginBottom","type"="number","defaultValue"=50,"label"="Bottom margin","possibleValues"=NULL),list("id"="marginLeft","type"="number","defaultValue"=30,"label"="Left margin","possibleValues"=NULL),list("id"="marginRight","type"="number","defaultValue"=15,"label"="Right margin","possibleValues"=NULL),list("id"="measurementGroupsAggregator","type"="categorical","defaultValue"="mean-stdev","label"="Aggregator for measurement groups","possibleValues"=list("mean-stdev","quartiles","count","min","max","sum")),list("id"="colLabel","type"="measurementsMetadata","defaultValue"="colLabel","label"="Color by","possibleValues"=NULL),list("id"="rowLabel","type"="measurementsAnnotation","defaultValue"="name","label"="Labels","possibleValues"=NULL),list("id"="offset","type"="categorical","defaultValue"="zero","label"="Offset","possibleValues"=list("zero","wiggle")),list("id"="interpolation","type"="categorical","defaultValue"="step-after","label"="Interpolation","possibleValues"=list("linear","step-before","step-after","basis","basis-open","basis-closed","bundle","cardinal","cardinal-open","monotone")),list("id"="scaleToPercent","type"="boolean","defaultValue"=TRUE,"label"="Scale to Percent","possibleValues"=NULL),list("id"="useGroupBy","type"="boolean","defaultValue"=FALSE,"label"="Use Group by","possibleValues"=NULL),list("id"="groupBy","type"="measurementsAnnotation","defaultValue"="name","label"="Group By","possibleValues"=NULL),list("id"="hoverOpacity","type"="number","defaultValue"=0.6,"label"="Hover Opacity","possibleValues"=NULL))

js_chart_colors = c("#393b79","#5254a3","#6b6ecf","#9c9ede","#637939","#8ca252","#b5cf6b","#cedb9c","#8c6d31","#bd9e39","#e7ba52","#e7cb94","#843c39","#ad494a","#d6616b","#e7969c","#7b4173","#a55194","#ce6dbd","#de9ed6")

app$chart_mgr$register_chart_type("StackedLinePlot", "epiviz.plugins.charts.StackedLinePlot", js_chart_settings=js_chart_settings, js_chart_colors=js_chart_colors)

app$server$start_server()

```
This opens a websocket connection between the interactive `R` session and the browser client. This will allow us to visualize data stored in the `Metaviz` server along with data in the interactive `R` session.

----
*Windows users:* In Windows platforms we need to use the `service` function to let the interactive `R` session connect to the `epiviz` web app and serve data requests. We then escape (using `ctl-c` or `esc` depending on your environment) to continue with the interactive `R` session. This is required anytime you want `metavizr` to serve data to the web app, for example, when interacting with the UI. (We are actively developing support for non-blocking sessions in Windows platforms).

```{r, eval=TRUE}
app$server$service()
```
----

### Adding Icicle plot to explore metagenomic Features

Once the browser is open we can visualize the metagenomic features from the `mouseData` object. We use the `plot` method to do so. 

```{r,eval=TRUE}
icicle_plot <- app$plot(mouseData, datasource_name="mouseData")
```

You should now see an icicle tree plot to explore the hierarchy of the metagenomic features from the `mouseData` object.

### Visualizing count matrix from the `mouseData` MRExperiment object

To get the count matrix from `mouseData`

```{r,eval=FALSE, echo=TRUE}
MRcounts(mouseData)
```

To visualize the count matrix, create a `metavizControl` object. The `metavizControl` object defines various settings like aggregation depth, normalization method, aggregate function etc. on the dataset.

```{r, eval=TRUE}
control = metavizControl(aggregateAtDepth=3,aggregateFun=function(x) colSums(x),
                          valuesAnnotationFuns=NULL,maxDepth=4,maxHistory=3,maxValue=NULL,
                          minValue=NULL,title="",n=10000,rankFun=sd,norm=TRUE,log=FALSE)
```

First, we'll create a heatmap visualization of the count matrix. For this lets choose all mouse samples whose diet is "Western"

```{r,eval=TRUE}
sampleIds = which(phenoData(mouseData)$diet == "Western")
```

We'll then use the `control` object to query for count matrix and plot a heatmap on the metaviz workspace.

```{r, eval=TRUE}

  # from control object
  norm=control$norm
  log = control$log
  n   = control$n
  rankFun = control$rankFun
  
  # query for counts
  mat = MRcounts(mouseData,norm=norm,log=log)
  
  # Since mouse data has > 10000 features choose a subset of features where normalized count values are > 5000
  otusToKeep = which(rowSums(mat) > 5000)
  n = min(length(otusToKeep),n)
  otuStats = apply(mat[otusToKeep,], 1, rankFun)
  otuIndices = otusToKeep[order(otuStats, decreasing = TRUE)[1:n]]
```

Now we have the OTU Matrix, lets add this data to metavizr as a new data source.

```{r, eval=TRUE}
  heatMapObj = app$data_mgr$add_measurements(mouseData[otuIndices,], datasource_name = "heatmapObj")
  measurements = heatMapObj$get_measurements()
  heatmap = app$chart_mgr$visualize("HeatmapPlot", measurements[sampleIds])
```

Using the same data, we can also visualize it in a stacked plot to see the abundance of various features across samples. Since the measurements are already added to metavizr, we only need to add a stacked line plot. 

```{r, eval=TRUE}
  stackedPlot = app$chart_mgr$visualize("StackedLinePlot", measurements[sampleIds])
```

### Close Metavizr and end session

To close the connection with metaviz and the R session, we use the stop_app function.

```{r,eval=TRUE}
  app$stop_app()
```

### SessionInfo

```{r session-info, cache=FALSE}
sessionInfo()
```


